---
title: "How good do you have to be to win a tennis match?"
author: "William Munn"
date: "2/15/2022"
output: 
  html_document:
    theme: united
---

```{r,include=FALSE}
# This is needed to set the directory for all subsequent chunks
knitr::opts_knit$set(root.dir = '/Users/williammunn/Documents/github/tennis/functions/')
```

```{r,include=FALSE}
library(data.table)
library(ggplot2)
source("simulate_match.R")
```

What is the relationship between the percentage of points a player will win on their serve, and the percentage of *games* that same player will win? The unique scoring system in tennis means that this relationship is not linear.

Without doing the maths, this relationship can be inferred by simulating games of tennis. The `play.game()` function I have written makes this pretty straightforward. This function returns a list of two objects. The first is simply the winner of the game. The second is a dataframe that records the outcome of each game:

```{r}
result <- play.game("P1",0.6) # ignoring optional arguments that allow a game to begin mid-way
result[[1]]
result[[2]]
```

Here you can see the results for a player who wins, on average, 60% of their points played on serve. The most important thing is the *on average* part. Since the outcomes of individual points rely on a random number being generated, it's possible (but unlikely) for a player who wins, say, 80% of their service points to still lose a simulated game.

How unlikely, though? If we do a sufficiently large enough number of simulations, we can take a look.

```{r}
ntimes <- 1000
winners <- sapply(1:ntimes,function(x) {play.game("P1",0.8)[[1]]})
sum(winners=="P1")/length(winners)
```

The estimate printed below of `r paste(round(100*(sum(winners=="P1")/length(winners)), 2), "%", sep="")` was calculated based on `r ntimes` games being played. Next question: how many times does the simulation need to be run in order for the results to be reliable? The chart below illustrates how quickly the result will converge:

```{r, fig.width = 3, fig.height = 3}
# record how the result changes with respect to number of simulatons
ntimes <- 2500
winners <- sapply(1:ntimes,function(x) {play.game("P1",0.8)[[1]]})
graphdata <- data.table(n = c(1:ntimes),winner = winners)[,pct:=cumsum(winner=="P1")/.I]
p <- ggplot(graphdata, aes(x=n, y=pct)) + geom_line(colour = "cyan 3")
p
```

A few thousand runs seems like enough, then, to give us a reliable answer: a player who wins 80% of points on their serve would expect to win around **98%** of their service games - essentially all of them.

At the beginning of this post, I said that the relationship between a player's percentage of points won on serve and their percentage of *games* won on serve was non-linear. We can show this with - you guessed it - more simulations.

```{r, cache = TRUE, fig.width = 3, fig.height = 3}
# relationship between points won and games won
ntimes <- 1000
srvpct <- rep(seq(0,1,0.025),ntimes)
results <- data.table(srv_pct = srvpct, winner = sapply(srvpct, function(x) {play.game("P1",x)[[1]]}))
results <- results[,.(win_pct = sum(winner == "P1")/.N),by = srv_pct]
p <- ggplot(results, aes(x=srv_pct, y=win_pct)) + geom_line(colour = "cyan 3", size = 0.8)
p
```

In order words: if you're serving, then the gain you get from increasing your percentage of points won on server is the greatest around 50%. Too much higher than that, the odds of you holding serve are a sure thing, anyway, so the gain becomes less.

How does a players strength on serve affect their probability of winning sets and matches? To understand this, we need an assumption about their opponent's strength on serve, too. Here are the chance's of winning the set, for a range of different opponents:

```{r, include = FALSE}
# re define some functions to speed them up
play.set <- function(arg.set.first.server,arg.set.p1.serve.pct,arg.set.p2.serve.pct) {
  # initial values
  set.over <- FALSE
  p1.games <- 0
  p2.games <- 0
  tiebreak <- FALSE
  current.server <- arg.set.first.server
  set.most.recent.server <- current.server
  # play games until the set is over
  while(set.over == FALSE) {
    if (!tiebreak) {
      # play a game
      game.object <- play.game(arg.game.server = current.server, arg.game.serve.pct = ifelse(current.server == 'P1', arg.set.p1.serve.pct, arg.set.p2.serve.pct))
      # immediately switch off the inherited switch, so that any future games start at 0-0
      inherit.game <- FALSE
      # retrieve the winner of the game
      game.winner <- game.object
      # update the most recent server variable
      set.most.recent.server <- current.server
      # award winner a game
      if (game.winner == 'P1') {
        p1.games <- p1.games + 1
      } else {
        p2.games <- p2.games + 1
      }
      # check if set is over
      if((p1.games == 6 & p2.games <= 4) | (p2.games == 6 & p1.games <= 4) | p1.games == 7 | p2.games == 7) {
        set.over <- TRUE
      }
      # swap server
      current.server = ifelse(current.server == 'P1','P2','P1')
      # check if tiebreak needed
      if (p1.games == 6 & p2.games == 6) {
        tiebreak <- TRUE
      }
    }
    if(tiebreak) {
      game.object <- play.tiebreak(arg.tiebreak.first.server = current.server,arg.tiebreak.p1.serve.pct = arg.set.p1.serve.pct,arg.tiebreak.p2.serve.pct = arg.set.p2.serve.pct)
      game.winner <- game.object
      set.over <- TRUE
      # award winner of tiebreak a game
      if (game.winner == 'P1') {
        p1.games <- p1.games + 1
      } else {
        p2.games <- p2.games + 1
      }
    }
  }
  if (p1.games > p2.games) {
    set.winner <- 'P1'
  } else {
    set.winner <- 'P2'
  }
  return(set.winner)
}



# define play.game
play.game <- function(arg.game.server,arg.game.serve.pct) {
  # initial values
  p1.points <- 0
  p2.points <- 0
  current.server <- arg.game.server
  game.over <- FALSE
  # play points until the game is over
  while(game.over == FALSE) {
    # play a point
    point.winner <- play.point(arg.point.server = current.server, arg.point.serve.pct = arg.game.serve.pct)
    # increment points of the winner of the point just played
    if (point.winner == 'P1') {
      p1.points <- p1.points + 1
    } else {
      p2.points <- p2.points + 1
    }
    if (p1.points == 4 & p2.points == 4) {
      p1.points <- 3
      p2.points <- 3
    }
    # check if the game is over
    if ((p1.points >= 4 | p2.points >= 4) &
        (abs(p1.points - p2.points) >= 2)) {
      game.over <- TRUE
    }
  }
  if (p1.points > p2.points) {
    game.winner <- 'P1'
  } else {
    game.winner <- 'P2'
  }
  return(game.winner)
}


play.tiebreak <- function(arg.tiebreak.first.server,arg.tiebreak.p1.serve.pct,arg.tiebreak.p2.serve.pct) {
  # initial values
  tiebreak.over <- FALSE
  tiebreak.current.server <- arg.tiebreak.first.server
  p1.points <- 0
  p2.points <- 0
  tiebreak.serves.remaining <- 1 # first person to serve only serves once
  # play the remainder of the tiebreak
  while(tiebreak.over == FALSE) {
    # play the current server's 2 points
    while (tiebreak.serves.remaining > 0) {
      # play the first point on this player's server
      point.winner <- play.point(arg.point.server = tiebreak.current.server, arg.point.serve.pct = ifelse(tiebreak.current.server == 'P1',arg.tiebreak.p1.serve.pct,arg.tiebreak.p2.serve.pct))
      # award point to the winner
      if (point.winner == 'P1') {
        p1.points <- p1.points + 1
      } else {
        p2.points <- p2.points + 1
      }
      # increment the points served by this current server
      tiebreak.serves.remaining <- tiebreak.serves.remaining - 1
      # check if tiebreak should be ended
      if ((p1.points >= 7 | p2.points >= 7) & (abs(p1.points - p2.points) > 1)) {
        tiebreak.over <- TRUE
        break
      }
    }
    # after the first point players get 2 serves each
    tiebreak.serves.remaining <- 2
    # swap server
    tiebreak.current.server <- ifelse(tiebreak.current.server == "P1","P2","P1")
  }
  # determine who wins the tiebreak
  if (p1.points > p2.points) {
    tiebreak.winner <- 'P1'
  } else {
    tiebreak.winner <- 'P2'
  }
  # return the winner of the tiebreak
  return(tiebreak.winner)
}
```

```{r, cache = TRUE, fig.width = 3, fig.height = 3}
# relationship between points won and sets won
simulations <- 100
p1srange <- rep(seq(0.1,1,0.025),simulations)
p2srange <- rep(seq(0.5,0.9,0.1),simulations)
comb <- expand.grid(p1 = p1srange, p2 = p2srange)
results <- apply(comb, 1, function(x) play.set("P1",x[1],x[2])[[1]])
results2 <- data.table(p1 =comb$p1, p2=comb$p2, winner=results)[,.(win_pct = sum(winner=="P1")/.N),by=.(p1,p2)][order(p1,p2)]
p <- ggplot(results2, aes(x=p1, y = win_pct, group = p2, color = p2)) + geom_line(size = 0.8) + theme(legend.position = "none")
p
```

The lowest line represents an opponent who wins, on average, 50% of their points on serve. The remaining lines show opponents who win 60%, 70%, 80% and 90% of points on their serve.







